---
title: "RNA-seq Analysis for GSE253154: MTR4 cKO vs WT in Mouse Spermatogonia"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Study Overview

This analysis examines RNA-seq data from GSE253154 (Zhang et al., *Nat Commun* 2025).
The study investigates the role of MTR4, a nuclear RNA exosome co-factor, in mouse
spermatogenesis. Sorted spermatogonia were compared between MTR4 conditional knockout
(cKO) mice and wild-type (WT) controls. The design is a single-factor experiment with
two groups of three biological replicates each (6 samples total), sequenced on the
Illumina NovaSeq 6000 platform.

---

## Load Libraries and Data

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
# NOTE: using class code from FGT RNA-seq tutorial (Simon Tomlinson, 2026)
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(pheatmap)
library(EnhancedVolcano)
library(stringr)
library(scatterplot3d)
library(ggplot2)
library(RColorBrewer)
library(biomaRt)

# Load the feature counts data
# Set working directory to study folder before running:
setwd("~/Functional_Genomics_Technologies/ICA/FGT_ICA_s2793337_GSE253154/GSE253154")
load("mytable_feaures")

ls()
dir()
```

---

## Process Sample Annotation

```{r input_processing, echo=FALSE}
# Extract count table and column names
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
table_rnaseq <- (mytable_feaures)$counts
colnames_rnaseq <- colnames(table_rnaseq)

# Build annotation data frame from column names
adf <- data.frame(Colnames = colnames_rnaseq)
locc <- str_locate_all(pattern = "Aligned.sortedByCoord.out.bam", adf$Colnames)
adf$ShortGEOName <- substr(adf$Colnames, 1, locc[[1]][, "start"] - 1)

# Sample annotation from GEO page for GSE253154:
# GSM8015124  Mtr4-WT-rep1
# GSM8015125  Mtr4-WT-rep2
# GSM8015126  Mtr4-cKO-rep1
# GSM8015127  Mtr4-cKO-rep2
# GSM8554711  Mtr4-WT-rep3
# GSM8554712  Mtr4-cKO-rep3
# NOTE: samples are NOT in sequential group order - group labels must be assigned explicitly

adf$ShortNames <- c("WT_1", "WT_2", "cKO_1", "cKO_2", "WT_3", "cKO_3")

# Extract treatment group from short names
split_names <- str_split(adf$ShortNames, pattern = "_")
treatment <- c()
for (i in 1:length(split_names)) {
  treatment <- c(treatment, split_names[[i]][1])
}
adf$Treatment <- treatment

# Assign colours per treatment group
z <- factor(adf$Treatment)
adf$Colours <- rainbow(length(levels(z)))[z]

# Verify column name matching
rownames(adf) <- adf$Colnames
if (identical(colnames_rnaseq, rownames(adf)) == FALSE) {
  print("Names do not match - check sample order!")
} else {
  print("Names match - OK to proceed")
}

# Switch to short names for figures
rownames(adf) <- adf$ShortNames
colnames(table_rnaseq) <- adf$ShortNames
colnames_rnaseq <- colnames(table_rnaseq)

print(adf[, c("ShortNames", "Treatment")])
```

---

## Load Data into DESeq2

```{r loading_deseq, echo=FALSE}
# Create DESeq2 object with Treatment as the design factor
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
dds_rnaseq <- DESeqDataSetFromMatrix(
  countData = table_rnaseq,
  colData = adf,
  design = ~ Treatment
)

dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
```

---

## Quality Filtering

```{r apply_filters, echo=FALSE}
# Remove genes with very low counts across all samples
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep, ]
cat("Number of genes retained after filtering:", nrow(dds_rnaseq), "\n")
```

---

## Normalisation

```{r normalisation, echo=FALSE}
# Estimate size factors for library-size normalisation
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)
```

---

## Visualise Normalised Data — Boxplots

```{r plot_normalised, echo=FALSE, fig.cap="Figure 1. Boxplots of gene expression under different normalisations. From left to right: log2 normalised counts, log FPM, VST, and rlog. Consistent distributions across samples indicate successful normalisation."}
# Generate multiple normalisations for comparison
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
counts_rnaseq <- log2(counts(dds_rnaseq, normalized = TRUE))
fpm_rnaseq <- log(fpm(dds_rnaseq))
vsd_rnaseq <- vst(dds_rnaseq, blind = TRUE)
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)

par(mfrow = c(2, 2), mar = c(7, 4, 3, 1))
boxplot(counts_rnaseq, las = 2, main = "Log2 Normalised Counts",
        ylab = "Log2 Counts", col = adf$Colours, cex.axis = 0.7)
boxplot(fpm_rnaseq, las = 2, main = "Log FPM",
        ylab = "Log FPM", col = adf$Colours, cex.axis = 0.7)
boxplot(assay(vsd_rnaseq), las = 2, main = "VST Normalised",
        ylab = "VST", col = adf$Colours, cex.axis = 0.7)
boxplot(assay(rld_rnaseq), las = 2, main = "rlog Normalised",
        ylab = "rlog", col = adf$Colours, cex.axis = 0.7)
par(mfrow = c(1, 1))
```

---

## PCA and Sample Distance Analysis

```{r more_explore, echo=FALSE, fig.cap="Figure 2. Principal Component Analysis (PCA) of rlog-normalised expression data. Each point represents one sample, coloured by treatment group (WT = one colour, cKO = other). Clustering by group indicates biological signal."}
# PCA using rlog normalised data
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale = TRUE)

# 3D PCA
s3d <- scatterplot3d(pca$x[, 1:3], pch = 19, color = adf$Colours,
                     main = "3D PCA Plot — MTR4 cKO vs WT",
                     xlab = "PC1", ylab = "PC2", zlab = "PC3")
s3d.coords <- s3d$xyz.convert(pca$x[, 1:3])
text(s3d.coords$x, s3d.coords$y,
     labels = colnames(rld_rnaseq), pos = 3, offset = 0.5, cex = 0.7)

# 2D PCA with ggplot2
pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Sample = rownames(pca$x),
  Treatment = adf$Treatment
)

pct_var <- round(100 * summary(pca)$importance[2, 1:2], 1)

ggplot(pca_df, aes(x = PC1, y = PC2, colour = Treatment, label = Sample)) +
  geom_point(size = 4) +
  geom_label(nudge_y = 1, size = 3) +
  xlab(paste0("PC1 (", pct_var[1], "% variance)")) +
  ylab(paste0("PC2 (", pct_var[2], "% variance)")) +
  ggtitle("PCA — MTR4 cKO vs WT Spermatogonia") +
  theme_bw()

# Sample distance matrix
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists
```

---

## Sample Distance Heatmap and DESeq2 PCA

```{r more_heatmaps, echo=FALSE, fig.cap="Figure 3. Sample-to-sample distance heatmap based on rlog expression values. Darker colours indicate more similar samples. Tight within-group clustering supports good replicate reproducibility."}
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
library(pheatmap)
library(RColorBrewer)

sampleDistMatrix <- as.matrix(sampleDists)
colours <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colours,
         main = "Sample-to-Sample Distance Matrix")

# Built-in DESeq2 PCA
ress <- plotPCA(rld_rnaseq, intgroup = c("Treatment"))
ress <- ress + geom_label(aes(label = name)) +
  ggtitle("DESeq2 PCA — MTR4 cKO vs WT") +
  theme_bw()
plot(ress)
```

---

## Differential Expression Analysis

```{r differential_expression, echo=FALSE, fig.cap="Figure 4. MA plot of differential expression (cKO vs WT). Each point is a gene; the y-axis shows log2 fold change and the x-axis shows mean expression. Red points are significantly differentially expressed (padj < 0.05)."}
# Run DESeq2 differential expression
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
dds_rnaseq <- DESeq(dds_rnaseq)

# Specify contrast: cKO vs WT
result_treatment <- results(dds_rnaseq, contrast = c("Treatment", "cKO", "WT"))

# MA plot
plotMA(result_treatment,
       main = "MA Plot — MTR4 cKO vs WT",
       ylab = "Log2 Fold Change (cKO / WT)",
       xlab = "Mean Normalised Counts",
       ylim = c(-8, 8))

# Summary of significant genes
cat("Genes with padj < 0.05:\n")
table(result_treatment$padj < 0.05)
cat("\nGenes with padj < 0.01:\n")
table(result_treatment$padj < 0.01)

# Subset significant results
result_treatment_selected <- subset(result_treatment, padj < 0.05)
cat("\nNumber of significant DE genes (padj < 0.05):", nrow(result_treatment_selected), "\n")

# Order by absolute test statistic (most confident DE genes first)
result_treatment <- result_treatment[order(abs(result_treatment$stat), decreasing = TRUE), ]
head(result_treatment, 10)
```

---

## Dispersion Plot and Cook's Distance Outlier Check
```{r dispersion_cooks, echo=FALSE, fig.cap="Figure 5. Top: DESeq2 dispersion estimates. Black dots show raw per-gene dispersion estimates, the red line shows the fitted trend, and blue dots show final shrunken estimates used for testing. Bottom: Cook's distance per sample. Genes where one sample has a disproportionately extreme count are flagged as outliers and assigned NA for p-value."}
# Dispersion plot - assesses how well DESeq2 modelled biological variability
# This is an extension beyond class materials - dispersion modelling is key to
# understanding why some genes receive NA p-values in low-count datasets
# Reference: Love et al. 2014, Genome Biology

plotDispEsts(dds_rnaseq,
             main = "DESeq2 Dispersion Estimates — MTR4 cKO vs WT",
             xlab = "Mean Normalised Counts",
             ylab = "Dispersion")

# Cook's distance - identifies influential outlier observations per gene
# Genes where one sample has an extreme count relative to others get flagged
# This explains the NA p-value for ENSMUSG00000057170 in our results
# Reference: Cook 1977; implemented in DESeq2 via results()

cooks <- assays(dds_rnaseq)[["cooks"]]

boxplot(log10(cooks + 1e-10),
        las = 2,
        col = adf$Colours,
        main = "Cook's Distance per Sample",
        ylab = "log10(Cook's Distance)",
        names = colnames(cooks))

# Report genes flagged as outliers
na_pval_genes <- rownames(result_treatment)[is.na(results(dds_rnaseq)$pvalue)]
cat("Number of genes with NA p-value (Cook's distance outliers):",
    length(na_pval_genes), "\n")
cat("Top outlier gene(s):\n")
print(na_pval_genes[1:min(5, length(na_pval_genes))])

# Show Cook's distances for the top outlier gene
if(length(na_pval_genes) > 0){
  cat("\nCook's distances for top outlier gene", na_pval_genes[1], ":\n")
  print(cooks[na_pval_genes[1], ])
}
```

---

## Heatmaps of Top Differentially Expressed Genes

```{r differential_expression_heatmaps, echo=FALSE, fig.cap="Figure 5. Heatmaps of top differentially expressed genes. Top: top 50 genes by test statistic, row-scaled rlog values. Bottom: all significant genes (padj < 0.05)."}
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
top50 <- rownames(result_treatment)[1:50]

pheatmap(assay(rld_rnaseq)[top50, ],
         scale = "row",
         show_rownames = FALSE,
         main = "Top 50 DE Genes — rlog Values, Row Scaled",
         annotation_col = data.frame(Treatment = adf$Treatment,
                                     row.names = rownames(adf)))

# All significant genes
sig_genes <- rownames(result_treatment_selected)
if (length(sig_genes) > 1) {
  pheatmap(assay(rld_rnaseq)[sig_genes, ],
           scale = "row",
           show_rownames = FALSE,
           main = "All Significant DE Genes (padj < 0.05) — Row Scaled",
           annotation_col = data.frame(Treatment = adf$Treatment,
                                       row.names = rownames(adf)))
}
```

---

## Gene Annotation via BiomaRt

```{r build_gene_annotation, echo=FALSE, message=FALSE, warning=FALSE}
# Download or load cached Ensembl annotation for mouse genes
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
if (file.exists("resultAnnot.RData")) {
  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
} else {
  # Try mirrors in order until one works - main Ensembl site can be unreliable
  ensembl_mirrors <- c(
    "https://useast.ensembl.org",
    "https://uswest.ensembl.org",
    "https://asia.ensembl.org",
    "https://www.ensembl.org"
  )
  
  mart <- NULL
  for(mirror in ensembl_mirrors){
    tryCatch({
      mart <- biomaRt::useDataset(
        dataset = "mmusculus_gene_ensembl",
        mart = useMart("ENSEMBL_MART_ENSEMBL", host = mirror)
      )
      cat("Connected successfully via:", mirror, "\n")
      break
    }, error = function(e){
      cat("Failed to connect via:", mirror, "\n")
    })
  }
  
  if(is.null(mart)) stop("Could not connect to any Ensembl mirror")
  
  resultAnnot <- biomaRt::getBM(
    values = rownames(dds_rnaseq),
    attributes = c("ensembl_gene_id", "external_gene_name",
                   "chromosome_name", "start_position",
                   "end_position", "description", "strand"),
    filters = "ensembl_gene_id",
    mart = mart
  )
  save(resultAnnot, file = "resultAnnot.RData")
  print("Downloaded and saved Ensembl annotation...")
}
```
---

## Merge Annotation with Results

```{r merge_gene_annotation, echo=FALSE}
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
names_annot <- resultAnnot[, 1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) <- names_annot

grr <- resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id), ]
resultAnnot <- grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))

nice_names <- paste(resultAnnot$ensembl_gene_id, resultAnnot$external_gene_name, sep = "_")
resultAnnot$nice_names <- nice_names

# Regenerate rlog after DESeq run
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)

# Heatmap with gene names for top 50
idx2 <- match(rownames(result_treatment)[1:50], rownames(dds_rnaseq))
plotme <- (rld_rnaseq)[rownames(result_treatment)[1:50], ]
rownames(plotme) <- resultAnnot$nice_names[idx2]

png(filename = "heatmap_top50_annotated.png", width = 900, height = 1100)
pheatmap(assay(plotme), scale = "row", fontsize_row = 8,
         cellheight = 12, cellwidth = 20,
         treeheight_row = 40, treeheight_col = 40,
         main = "Top 50 DE Genes — Annotated")
dev.off()

pheatmap(assay(plotme), scale = "row", fontsize_row = 8,
         cellheight = 12, cellwidth = 20,
         treeheight_row = 40, treeheight_col = 40,
         main = "Top 50 DE Genes — Annotated",
         annotation_col = data.frame(Treatment = adf$Treatment,
                                     row.names = rownames(adf)))
```

---

## Gene Biotype Analysis
```{r biotype_analysis, echo=FALSE, fig.cap="Figure 8. Gene biotype analysis. Left: biotype composition of all detected genes. Right: biotype composition of top 200 DE genes ranked by |Wald statistic|. Enrichment of specific biotypes in the DE list relative to background connects to MTR4's role as a nuclear RNA surveillance factor regulating non-coding and retrotransposon RNAs."}
# Gene biotype analysis - extension beyond class materials
# MTR4 is known to regulate non-coding RNAs and retrotransposon RNAs
# Checking whether DE genes are enriched for specific biotypes directly
# connects the analysis to the paper's central biological claims
# Reference: Zhang et al. 2025 Nat Commun
# Git checkpoint: git commit -m "Gene biotype analysis complete"

if(file.exists("resultAnnot_biotype.RData")){
  load("resultAnnot_biotype.RData")
  cat("Loaded cached biotype annotation...\n")
} else {
  ensembl_mirrors <- c(
    "https://useast.ensembl.org",
    "https://uswest.ensembl.org",
    "https://asia.ensembl.org",
    "https://www.ensembl.org"
  )
  mart_bio <- NULL
  for(mirror in ensembl_mirrors){
    tryCatch({
      mart_bio <- biomaRt::useDataset(
        dataset = "mmusculus_gene_ensembl",
        mart = useMart("ENSEMBL_MART_ENSEMBL", host = mirror)
      )
      cat("Connected via:", mirror, "\n")
      break
    }, error = function(e) cat("Failed:", mirror, "\n"))
  }
  biotype_annot <- biomaRt::getBM(
    values = rownames(dds_rnaseq),
    attributes = c("ensembl_gene_id", "gene_biotype"),
    filters = "ensembl_gene_id",
    mart = mart_bio
  )
  save(biotype_annot, file = "resultAnnot_biotype.RData")
  cat("Downloaded and saved biotype annotation...\n")
}

# Biotype of all detected genes (background)
all_biotypes <- sort(table(biotype_annot$gene_biotype), decreasing = TRUE)
top_biotypes <- names(all_biotypes)[1:10]

# Collapse rare biotypes into "other"
biotype_annot$biotype_collapsed <- ifelse(
  biotype_annot$gene_biotype %in% top_biotypes,
  biotype_annot$gene_biotype, "other"
)

# Biotype of top 200 DE genes
top200_ids <- rownames(result_treatment)[1:200]
top200_biotypes <- biotype_annot[biotype_annot$ensembl_gene_id %in% top200_ids, ]
top200_biotypes$biotype_collapsed <- ifelse(
  top200_biotypes$gene_biotype %in% top_biotypes,
  top200_biotypes$gene_biotype, "other"
)

# Plot comparison
par(mfrow = c(1, 2), mar = c(10, 4, 3, 1))

all_collapsed <- sort(table(biotype_annot$biotype_collapsed), decreasing = TRUE)
barplot(all_collapsed, las = 2, col = "steelblue",
        main = "All Detected Genes",
        ylab = "Number of genes", cex.names = 0.65)

top200_collapsed <- sort(table(top200_biotypes$biotype_collapsed), decreasing = TRUE)
barplot(top200_collapsed, las = 2, col = "coral",
        main = "Top 200 DE Genes",
        ylab = "Number of genes", cex.names = 0.65)

par(mfrow = c(1, 1))

# Percentage breakdown for report
cat("\nBiotype breakdown — All detected genes (%):\n")
print(round(100 * all_collapsed / sum(all_collapsed), 2))
cat("\nBiotype breakdown — Top 200 DE genes (%):\n")
print(round(100 * top200_collapsed / sum(top200_collapsed), 2))

# Specifically check for retrotransposon-related genes in DE list
retro_types <- c("LTR_retrotransposon", "LINE_element",
                 "SINE_element", "retrotransposon")
retro_in_top200 <- top200_biotypes[top200_biotypes$gene_biotype %in% retro_types, ]
cat("\nRetrotransposon-related genes in top 200 DE genes:", nrow(retro_in_top200), "\n")
if(nrow(retro_in_top200) > 0) print(retro_in_top200)
```

---

## Annotated Results Table — Top 10 Differentially Expressed Genes

```{r top10_table, echo=FALSE}
# Add gene names to results table and display top 10
# Ordering by absolute test statistic (Wald stat) - this ranks genes by statistical
# confidence in differential expression, independent of fold change direction.
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)

idx2 <- match(rownames(result_treatment), rownames(resultAnnot))

# Some genes may not have annotation - replace NA nice_names with Ensembl ID
nice_names_safe <- resultAnnot$nice_names[idx2]
na_idx <- is.na(nice_names_safe)
nice_names_safe[na_idx] <- rownames(result_treatment)[na_idx]

# Check for any remaining duplicates and make unique
nice_names_safe <- make.unique(nice_names_safe)

result_treatment_annotated <- result_treatment
rownames(result_treatment_annotated) <- nice_names_safe

cat("Top 10 differentially expressed genes (ranked by |Wald statistic|):\n")
print(as.data.frame(result_treatment_annotated[1:10, ]))

# Also save the full results for GSEA input
result_for_gsea <- as.data.frame(result_treatment)
write.csv(result_for_gsea, file = "result_GSE253154.csv")
cat("\nFull results saved to result_GSE253154.csv for GSEA analysis\n")
```
---

## Volcano Plot

```{r make_volcano, echo=FALSE, fig.width=10, fig.height=10, fig.cap="Figure 6. Volcano plot of differential expression results (MTR4 cKO vs WT). X-axis: log2 fold change; Y-axis: -log10 adjusted p-value. Dashed lines indicate thresholds of padj = 1e-5 and |log2FC| = 2. Labelled genes are the top 15 most significant."}
# Add gene names to results for volcano plot
# NOTE: adapted from class code (FGT tutorial, Simon Tomlinson 2026)
result_volcano <- as.data.frame(result_treatment)
result_volcano$gene <- resultAnnot$external_gene_name[idx2]

# Replace empty or NA gene names with Ensembl ID
result_volcano$gene[is.na(result_volcano$gene) | result_volcano$gene == ""] <- 
  rownames(result_volcano)[is.na(result_volcano$gene) | result_volcano$gene == ""]

# Make gene names unique to avoid duplicate rowname error
result_volcano$gene <- make.unique(result_volcano$gene)
rownames(result_volcano) <- result_volcano$gene

# Select top genes to label — top 15 by adjusted p-value
top_label_genes <- head(
  rownames(result_volcano[order(result_volcano$padj, na.last = TRUE), ]),
  15
)

volcano_plot <- EnhancedVolcano(
  result_volcano,
  lab = rownames(result_volcano),
  x = "log2FoldChange",
  y = "padj",
  selectLab = top_label_genes,
  xlab = bquote(~Log[2] ~ "fold change (cKO / WT)"),
  ylab = bquote(~-Log[10] ~ "adjusted p-value"),
  title = "MTR4 cKO vs WT — Differential Expression",
  subtitle = "GSE253154 | Spermatogonia | Mouse",
  pCutoff = 1e-5,
  FCcutoff = 2.0,
  pointSize = 3.0,
  labSize = 5.0,
  labCol = "black",
  labFace = "bold",
  boxedLabels = TRUE,
  colAlpha = 4/5,
  legendPosition = "right",
  legendLabSize = 10,
  legendIconSize = 4.0,
  drawConnectors = TRUE,
  widthConnectors = 0.8,
  colConnectors = "black"
)

png("volcano_GSE253154.png", width = 900, height = 900)
plot(volcano_plot)
dev.off()

plot(volcano_plot)
```
---

## Session Information

```{r session_info, echo=FALSE}
sessionInfo()
```
